---
title: "DATA 643 Recommender Systems: Project 1 - Global Baseline Predictors and RMSE "
author: "Sarah Wigodsky"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---

The system recommends movies to viewers.  I asked 7 people to rate 9 movies and will make recommendations based on their reviews and try to predict their reviews.

```{r load-libraries, echo=FALSE}
suppressWarnings(suppressMessages(library(knitr)))
```


```{r input-data, echo=FALSE}
#inputing movie reviews
The_Help <- c(5,3,3,4,3,3, NA)
Solo <- c(4,4,5,5,NA,NA, 4)
Dr_Strangelove <- c(2,5,1,NA,NA,2,NA)
Fantastic_Beasts <- c(3,3,3,3,NA,NA,5)
Clue <- c(3,5,5,NA,NA,NA,5)
Isle_of_Dogs <- c(NA,1,NA,5,5,NA,NA)
Groundhog_Day <- c(5,5,3,4,4,4,5)
Princess_Bride <- c(4,4,4,5,NA,NA,5)
Goonies <- c(5,3,3,4,NA,NA,5)

reviews <- matrix(c(The_Help,Solo,Dr_Strangelove,Fantastic_Beasts,Clue,Isle_of_Dogs,Groundhog_Day,Princess_Bride,Goonies), nrow=7,ncol=9)
colnames(reviews) <- c("The_Help","Solo","Dr_Strangelove","Fantastic_Beasts","Clue","Isle_of_Dogs","Groundhog_Day","Princess_Bride","Goonies")
rownames(reviews) <- c("Sarah","Nadav","Alon","Deena","Robin","David","Dan")
```

The following table lists the names of the movies and the reviewers.  The reviews ranged from 1 to 5, where 1 is the lowest value and 5 is the greatest value.  Not every person has seen every movie, so there are some missing values.

```{r printing-review-table, echo=FALSE}
kable(reviews)
```

In order to create a predictor model, the data must be separated into a training set and a testing set.  Seventy percent of the data will be used for training and 30% of the data will be used for testing. The data taken for testing and training is assigned randomly.

```{r splitting-data, echo=FALSE}

set.seed(1)
entry_num <- length(reviews)
training_num <- floor(.7*entry_num)
random_nums <- sample(1:entry_num,training_num)

train_reviews = reviews
test_reviews = reviews
for (i in 1:entry_num){
  if (i %in% random_nums) 
    test_reviews[i]=NA
  else
    train_reviews[i]=NA
}
```  

The test data is:
```{r printing-test-review-table, echo= FALSE}
kable(test_reviews)
```  

The training data to build the recommender system is:
```{r printing-training-review-table, echo= FALSE}
kable(train_reviews)
``` 

To begin creating a system that can recommend a movie, the mean of the overall movie recommendations is found from the training set.
```{r calc-mean, echo=FALSE}
train_mean <- mean(train_reviews, na.rm=TRUE)
train_mean
```


```{r RMSE-function, echo=FALSE}
rmse <- function(df, mean_val){
  diff <- df-mean_val
  sq_diff <- diff^2
  mean_val <- mean(sq_diff, na.rm=TRUE)
  rmse <-sqrt(mean_val)
  return(rmse)
}  
```

The RMSE (root mean square error) between the mean value and each rating in the training data
```{r rmse-train, echo=FALSE}
rmse_train <- rmse(train_reviews,train_mean)
rmse_train
```

The RMSE (root mean square error) between the mean value and each rating in the test data
```{r rmse-test, echo=FALSE}
rmse_test <- rmse(test_reviews,train_mean)
rmse_test
```

Calculate the Bias for Each Reviewer using the Training Data by Subtracting the Mean Rating from the Mean User Rating
```{r bias-reviewer, echo=FALSE}
user_bias_df <-data.frame(stringsAsFactors = FALSE)

for (user in 1:nrow(train_reviews)){
  mean_row <- mean(train_reviews[user,], na.rm=TRUE)
  bias <- mean_row - train_mean 
  user_bias_df <- rbind(user_bias_df, list(row.names(train_reviews)[user], bias), stringsAsFactors=FALSE)
}
colnames(user_bias_df) <- c("user_name", "bias")
kable(user_bias_df)
```

Calculate the Bias for Each Movie using the Training Data by Subtracting the Mean Rating from the Mean Movie Rating

```{r bias-movie, echo=FALSE}
movie_bias_df <-data.frame(stringsAsFactors = FALSE)

for (movie in 1:ncol(train_reviews)){
  mean_col <- mean(train_reviews[,movie], na.rm=TRUE)
  bias <- mean_col - train_mean 
  movie_bias_df <- rbind(movie_bias_df, list(colnames(train_reviews)[movie],bias), stringsAsFactors=FALSE)
  
}
colnames(movie_bias_df) <- c("movie_name", "bias")
kable(movie_bias_df)
```

Calculate Baseline Predictors for Each User-Movie Combination by Adding the User Bias and Movie Bias to the Mean Rating

```{r baseline-predictor, echo=FALSE}
baseline_predictor <- reviews
for (col_num in 1:ncol(baseline_predictor)){
  movie_bias <- movie_bias_df$bias[col_num]
  for (row_num in 1:nrow(baseline_predictor)){
    user_bias <- user_bias_df$bias[row_num]
    baseline_predictor[row_num,col_num] = train_mean + user_bias + movie_bias 
  }
}
kable(baseline_predictor)
```

The RMSE (root mean square error) between the baseline predictor and each rating in the training set - 
```{r rmse-training-set, echo=FALSE}
rmse_train_baseline <- rmse(train_reviews,baseline_predictor)
rmse_train_baseline
```
The RMSE (root mean square error) between the baseline predictor and each rating in the test set -
```{r rmse-test-set, echo=FALSE}
rmse_test_baseline <- rmse(test_reviews,baseline_predictor)
rmse_test_baseline
```

Calculate the Percent Improvement in the RMSE for the Training Set
```{r, improvement-training, echo=FALSE}
100*(1-rmse_train_baseline/rmse_train)
```

Calculate the Percent Improvement in the RMSE for the Test Set
```{r, improvement-test, echo=FALSE}
100*(1-rmse_test_baseline/rmse_test)
```

###Summary:
I built a simple predictor system to predict a user's rating of a movie. The movie review data set was broken up into a training set and a testing set.  A crude predictor system entailed predicting each rating to be the mean of the user ratings for movies in the training set.  The root mean square error for the training set is 1.08 and the root mean square error for the testing set is 1.20.  This means that the average error between the predicted review and the actual review is a little over 1, which is fairly signficant since the rating system only ranges between 1 to 5.
\n
The bias was then calculated for each reviewer and each movie by subtracting the mean rating from the mean rating for each reviewer and movie.  A baseline predictor for each movie for each reviewer was then built by adding the mean rating to the mean for each movie and reviewer.  The baseline predictions were a 29% improvement over the prediction based simply on the mean of the data for the training set.  However the baseline predictor produced predictions that were about 21% worse than the prediction based simply on the mean for the testing set.
\n
This is not a successful predictor model.  I think a better predictor model could be built if there were more reviewers.  When 30% of the data was removed to create a testing set, there was very little data left to build a useful predictor model. 
